---
title: "AlphaFold Analysis"
author: "Keilyn Duarte (PID A16881868)"
format: html
---

Here we analyze our AlphaFold structure prediction models. The input directory/folder comes from the ColabFold server:

## 8. Custom analysis of resulting models

```{r}
# Change this for YOUR results dir name
results_dir <- "hivpdb_94b5b" 

```

```{r}
# File names for all PDB models
pdb_files <- list.files(path=results_dir,
                        pattern="*.pdb",
                        full.names = TRUE)

# Print our PDB file names
basename(pdb_files)
```

I will use the Bio3D package for analysis
```{r}
library(bio3d)
```

Align and superpose

```{r}
pdbs <- pdbaln(pdb_files, fit=TRUE, exefile="replace w msa file name")
pdbs
```

```{r}
pc <- pca(pdbs)
plot(pc)
```

## RMSD analysis
RMSD is a common measure of structural distance used in structural biology.

```{r}
rd <- rmsd(pdbs, fit=T)
```

```{r}
range(rd)
```

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```



```{r}
pc <- pca(pdbs)
plot(pc)
```

## RMSD Analysis
RMSD is a common measure of structural distance used in structural biology.

```{r}
library(pheatmap)

colnames(rd) <- paste0("m",1:5)
rownames(rd) <- paste0("m",1:5)
pheatmap(rd)
```

