---
title: "Lab 12"
author: "Keilyn Duarte (PID 16881868)"
format: pdf
---

## Section 4: Population Scale Analysis [HOMEWORK]

# Reading the expression genotype results file

```{r}
expresults <- read.table(file="rs8067378_ENSG00000172057.6.txt")

head(expresults)
```

> Q13. Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.

Find number of G/G genotype samples
```{r}
library(dplyr)

gg <- expresults %>%
  filter(geno == "G/G")

nrow(gg)

```

Find number of A/G genotype samples
```{r}
ag <- expresults %>%
  filter(geno == "A/G")

nrow(ag)
```

Find number of A/A genotype samples
```{r}
aa <- expresults %>%
  filter(geno == "A/A")

nrow(aa)
```

Find the median expression level for G/G genotype
```{r}
median(gg$exp)
```

Find the median expression level for the A/G genotype
```{r}
median(ag$exp)
```

Find the median expression level for the A/A genotype
```{r}
median(aa$exp)
```

> Q14. Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)

ggplot(expresults, aes(x=geno, y=exp)) +
  geom_boxplot() +
  xlab("Genotype") +
  ylab("Expression Level") +
  labs(title="Genotype vs. Expression level of ORMDL3")
  
```

The "A/A" genotype has the highest expression level of ORMDL3 and the "G/G" genotype has the lowest. This suggests that the SNP does indeed effect the expression level of ORMDL3.
