---
title: "Class 18: Pertussis Mini Project"
author: "Keilyn Duarte (PID A16881868)"
format: pdf
---

Pertussis (aka whooping cough) is a deadly lung infection caused by the bacteria B. Pertussis.

The CDC tracks Pertussis cases around the US. http://tinyurl.com/pertussiscdc

We can "scope" this data using the R **datapasta** package.

```{r, include=FALSE}
cdc <- data.frame(data.frame(
  year = c(1922L,1923L,1924L,
           1925L,1926L,1927L,1928L,
           1929L,1930L,1931L,
           1932L,1933L,1934L,1935L,
           1936L,1937L,1938L,1939L,
           1940L,1941L,1942L,
           1943L,1944L,1945L,1946L,
           1947L,1948L,1949L,1950L,
           1951L,1952L,1953L,1954L,
           1955L,1956L,1957L,
           1958L,1959L,1960L,1961L,
           1962L,1963L,1964L,1965L,
           1966L,1967L,1968L,
           1969L,1970L,1971L,1972L,
           1973L,1974L,1975L,1976L,
           1977L,1978L,1979L,1980L,
           1981L,1982L,1983L,
           1984L,1985L,1986L,1987L,
           1988L,1989L,1990L,1991L,
           1992L,1993L,1994L,
           1995L,1996L,1997L,1998L,
           1999L,2000L,2001L,2002L,
           2003L,2004L,2005L,2006L,
           2007L,2008L,2009L,
           2010L,2011L,2012L,2013L,
           2014L,2015L,2016L,2017L,
           2018L,2019L,2020L,
           2021L,2022L, 2024L),
  cases = c(107473,164191,
            165418,152003,202210,181411,
            161799,197371,166914,
            172559,215343,179135,
            265269,180518,147237,
            214652,227319,103188,183866,
            222202,191383,191890,
            109873,133792,109860,
            156517,74715,69479,120718,
            68687,45030,37129,
            60886,62786,31732,28295,
            32148,40005,14809,11468,
            17749,17135,13005,6799,
            7717,9718,4810,3285,
            4249,3036,3287,1759,
            2402,1738,1010,2177,2063,
            1623,1730,1248,1895,
            2463,2276,3589,4195,
            2823,3450,4157,4570,2719,
            4083,6586,4617,5137,
            7796,6564,7405,7298,
            7867,7580,9771,11647,
            25827,25616,15632,10454,
            13278,16858,27550,18719,
            48277,28639,32971,20762,
            17972,18975,15609,
            18617,6124,2116,3044, 35493)
))
```

```{r}
head(cdc)
```

> Q1 and Q2

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(year, cases) +
  geom_line() +
  geom_vline(xintercept=1946, col="blue") +
  geom_vline(xintercept=1996, col="red") +
  geom_vline(xintercept=2020, col="gray") +
  geom_vline(xintercept=2004, col="green")
```

> Q3. 

There were high case numbers before the first wP (whole-cell) vaccine roll out in 1946. Then, there was a rapid decline in case numbers until 2004, when we had our first large-scale outbreak of Pertussis again. There is also a notable COVID related dip and recent rapid rise.

> Q. What is different about the immune response to infection if you had an older wP vaccine vs the newer aP vaccine?

## Computational Models of Immunity - Pertussis Boost (CMI-PB)

The CMI-PB project aims to address this key question: what is different between aP and wP individuals?

We can get all of the data from this ongoing project via JSON API calls. 
For this, we will use the **jsonlite** package using `install.packages("jsonlite")`

```{r}
library(jsonlite)
subject <- read_json("http://cmi-pb.org/api/v5_1/subject", simplifyVector = TRUE)

head(subject)
```

> Q. How many individuals "subjects" are in this dataset?

```{r}
nrow(subject)
```

> Q4. How many wP and aP primmed individuals are in this dataset?

```{r}
table(subject$infancy_vac)
```

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

This is not representative of the US population but it is the bigg

Obtain more data from CMI-PB:

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector=T)

ab_data <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector=T)

head(specimen)
head(ab_data)
```

I now have 3 tables of data from CBI-PB: `subject`, `specimen`, and `ab_data`. I need to "join" these tables so I will have all the info I need to work with.

For this, we will us the `inner_join` function from the **dplyr** package.

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```

```{r}
dim(subject)
dim(specimen)
dim(meta)
```

Now we can join our `ab_data` table to `meta` so we can use the antibody level data.

```{r}
abdata <- inner_join(meta, ab_data)
head(abdata)
```

> Q. How many different antibody isotypes are there in this dataset?

```{r}
unique(abdata$isotype)
```

```{r}
table(abdata$isotype)
```

> Q. How many different antigens are there?

```{r}
table(abdata$antigen)
```

I want a plot of antigen levels across the whole dataset.

```{r}
ggplot(abdata) +
  aes(MFI, antigen) +
  geom_boxplot()
```


```{r}
ggplot(abdata) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

Antigens like FIM2/3, PT, and FELD1 have a large range of values. Others like Measles don't show much activity.

> Q. Are there differences at this whole-dataset level between aP and wP?

```{r}
ggplot(abdata) +
  aes(MFI_normalised, antigen, col=abdata$infancy_vac) +
  geom_boxplot()
```

```{r}
ggplot(abdata) +
  aes(MFI_normalised, antigen, col=abdata$infancy_vac) +
  geom_boxplot() +
  facet_wrap(~infancy_vac)
```

## Examine IgG Ab titer levels

For this, I need to select out just isotype IgG.

```{r}
igg <- abdata |>
  filter(isotype == "IgG")

head(igg)
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```

Digging in further to look at the time course of IgG isotype PT antigen levels across aP and wP individuals.

```{r}
## Filter to include 2021 data only
abdata.21 <- abdata |> filter(dataset == "2021_dataset")

## Filter to look at IgG PT data only
pt.igg <- abdata.21 |> 
  filter(isotype == "IgG",  antigen == "PT")
  
## Plot and color by infancy_vac (wP vs. aP)
ggplot(pt.igg) +
  aes(x=planned_day_relative_to_boost,
      y=MFI_normalised,
      col=infancy_vac,
      group=subject_id) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept=0, linetype="dashed") +
  geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

















